---
title: "Example Markdown Table 1 with gtsummary"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
# this code just sets up the document - dont change anything here

# setup options
# hide code
knitr::opts_chunk$set(echo = FALSE)
# hide warnings
knitr::opts_chunk$set(warning = FALSE)
#hide messages
knitr::opts_chunk$set(message = FALSE)
# continue to knit even if errors
# errors will be shown in output document
knitr::opts_chunk$set(error = TRUE)
```

This Markdown will generate example "table 1"s . It can be modified to suit your data. More information can be found here: https://www.danieldsjoberg.com/gtsummary/ 

```{r}
# Call the packages you need
library(tidyverse) # general data wrangling functions in this package
library(data.table) # data table objects tend to work nicer with the gtsummary package for some reason
library(Hmisc) # labeling functions from this package
library(gtsummary) # makes pretty tables
```

```{r}
#read in your dataset
df <- read.csv("example_df.csv")
df <- data.table(df) #makes df a "data table" object - works nicer with the gtsummary package
```

```{r}
## Variable Set Up Example

# Continuous Variables 
# format is: label(data_set$variable_name) <- "label you'd like for variable_name"

#age
label(df$age)<- "Age"
#height
label(df$height) <- "Height (cm)"
#weight
label(df$weight) <- "Weight (kg)"

#factor variables

#gender
df$gender <- factor(df$gender, levels = c("0", "1"))   # factor your existing varialbe
levels(df$gender) <- c("Male", "Female")               # label the levels of your factor
label(df$gender) <- "Gender"                           # label the variable

#education
df$education <- factor(df$education, levels = c("0", "1","2"))
levels(df$education) <- c("Less than High School", "High School", "College or More")
label(df$education) <- "Highest Education Level Attained"
```

## Table 1: Simple Summary Table

Put any description of table 1 you'd like here. This table summarizes all data and does not include any stratification or statistical tests.

```{r}
# select variables to make a table 
table_variaibles <- df %>%
  dplyr::select(
    age,
    height,
    weight,
    gender,
    education,
  )

# make the table
table_variaibles %>% tbl_summary(statistic = list(
                                    all_continuous() ~ "{mean} ({sd})",    # specify statistics for continuous variables
                                    all_categorical() ~ "{n} ({p}%)"       # specify statistics for categorical variables
                                  ),
                                  digits = all_continuous() ~ 2,           # change number of decimal places for continuous variables 
                                  missing_text = "(Missing)") %>%          # how do you want "missing" observarions to appear
                    modify_header(label ~ "**Variable**") %>%              # change text for top of variables column here
                    modify_caption("**Table 1: Descriptive Statistics**")  # change the table title here
```

## Table 2: Stratified Summary Table

Summary table stratified by a variable (in this case gender). Statistical tests can be performed by specifying "test = list(...)". You can exclude the "test = list(...)" code to use the gtsummary default tests. "add_overall(last=TRUE)" adds a "total" column as the last column in the table. Setting it to =TRUE makes the "total" column appear first in the table. 

```{r}
# this code can be used to generate a table split by a variable like gender

table_variaibles %>% tbl_summary(by = gender,                                # specify the variable you'd like to split the table by
                                 statistic = list(
                                    all_continuous() ~ "{mean} ({sd})",
                                    all_categorical() ~ "{n} ({p}%)"
                                  ),
                                  digits = all_continuous() ~ 2,
                                  missing_text = "(Missing)") %>%
                    add_p(test = list(
                                    all_continuous() ~ "t.test",             # specify test type for continuous variables
                                    all_categorical() ~ "chisq.test"),       # specify test type for categorical variables
                      pvalue_fun = ~ style_pvalue(.x, digits = 2)) %>%       # add p-values and specify decimal places
                    add_overall(last = TRUE) %>%                             # add a "total" column to the table
                    modify_header(label ~ "**Variable**") %>%
                    modify_spanning_header(c("stat_1", "stat_2") ~ "**Gender**") %>%    # a header to span the "gender" columns
                    modify_caption("**Table 2. Demograhpics by Gender**") %>%
                    separate_p_footnotes() %>%                               # add footnotes for statistical tests
                    bold_labels()                                            # make the labels bold
```

## Table 3: Linear Regression Model

This data was randomly generated so model results are not what you'd expect necessarily. 

```{r}
mod1<-lm(weight~age+gender+height, data = df)                          # make a regression model

tbl1<-tbl_regression(mod1, intercept = TRUE) %>%                       # specify the intercept should be included in the table
                      modify_caption("**Regression model of Weight**") # table title
tbl1                                                                   # print the table

```

## Table 4: Combine Multiple Linear Regression Models into one table

This data was randomly generated so model results are not what you'd expect necessarily. 

```{r}
mod2<-lm(height~age+gender+weight, data = df)                           #make a second regression model

tbl1<-tbl_regression(mod1, intercept = TRUE) %>%                        # Re-make 1st table so we can change the title
                      modify_caption("**Regression Models**")           # change table title
tbl2<-tbl_regression(mod2, intercept = TRUE)                            #make a second table

tbl_merge(list(tbl1,tbl2),
          tab_spanner = c("**Weight**", "**Height**"))

```